\textbf{ Discovery of a disrupting open cluster far into the Milky Way halo:\\ a recent star formation event in the leading arm of the Magellanic stream?}

### 2018-11-14 ms.tex
apw edits

We report the discovery of a young ($\tau \sim \clage$), low-mass ($M \sim 1200~\msun$), metal-poor ($\feh \sim \clfeh$) stellar association in the Milky Way halo: \clustername.
From isochrone fitting to follow-up \decam\ imaging of the cluster, we also measure its distance, $D \approx \cldist$.
At its present Galactocentric position $(R, z) \sim (23, 15)~\textrm{kpc}$, the association is (on the sky) near the leading arm of the gas stream emanating from the Magellanic cloud system, but is located $\approx 60^\circ$ from the Large Magellanic Cloud (LMC) center on the other side of the Milky Way disk.
If we assume that the cluster is co-located with \hi gas in the stream, we directly measure the distance to the leading arm of the Magellanic stream.
At this location relative to the LMC, the measured distance is inconsistent with predictions from models of the LMC/SMC interaction and infall into the Milky Way that do not account for ram pressure and gas interaction with Milky Way disk.
The estimated age of \clustername\ is consistent with the time of last passage of the leading arm gas through the Galactic midplane, and we therefore speculate that this star-formation event was triggered by the last disk midplane passage, which occurred at a Galactocentric radius $R \approx 30~\kpc$.
However, most details of this idea remain a mystery: the Magellanic stream has low column density, the Milky Way disk at this large radius has low gas density, and the relative velocity of the leading arm gas and Milky Way gas is large. However it formed, the discovery of a young stellar cluster in the Milky Way halo presents an interesting opportunity for study.
\end{abstract}

@@

We then query all objects from the \gaia\ \DR{2} catalog within a rectangle centered on the nominal position of this group, with a width of $5^\circ$ and a height of $5^\circ$ in the equatorial (ICRS) coordinate system (see Appendix~\ref{sec:queries} for the database query).
\figurename~\ref{fig:cmds} shows the \gaia\ data for this region:
The left panel shows the \gaia\ color-magnitude diagram, extinction corrected following the procedure used in \citet{Danielski:2018}, with the blue over-density highlighted by the polygon (blue) and under-plotted with a $100~\textrm{Myr}$, $\feh = -1.1$ \acronym{MIST} isochrone \citep[red line;][]{Dotter:2016, Choi:2016, Paxton:2011, Paxton:2013, Paxton:2015}.
Surprisingly, this group appears to be a young, distant main sequence, rather than an old population of horizontal branch stars.
The middle and right panels of \figurename~\ref{fig:cmds} show sky positions and proper motions, respectively, of all stars in this sky region (grey background density), and only stars in the CMD selection polygon as black markers.

The \gaia\ data demonstrate the presence of a young, distant, spatially-clustered, and co-moving stellar over-density --- named \clustername\ --- but the \gaia\ photometry is too shallow to resolve anything but the brightest main sequence stars.
The spatial morphology of the cluster is large on the sky, and interestingly substructured with at least two subcomponents (labeled a and b in \figurename~\ref{fig:cmds}) that are indistinguishable in terms of their proper motions and uncertainties.
With the \gaia\ data alone, the distance, age, and metallicity of the cluster therefore cannot be determined, as these quantities are degenerate where the main sequence is nearly vertical.
In the next section, we describe deeper \decam\ imaging obtained over a portion of the cluster.

@@

\figurename~\ref{fig:decam-field} shows the sky positions of the \gaia\ CMD-selected sources (black markers; see same in middle panel, \figurename~\ref{fig:cmds}), along with point sources identified in the $g$-band data obtained with \decam: sources in control fields are shown as red markers, and sources in cluster fields are shown as blue markers.
We note that the smaller over-density located northeast of the \decam\ field was not followed up in this work.

@@

We measure the mean proper motion of \clustername\ by constructing a probabilistic model of the cluster and background populations using astrometric data from \gaia.
We start by selecting all stars with $(\bprp)_0 < 0.35$ to remove low-mass and old main sequence star contamination in the region.
\figurename~\ref{fig:pm-members}, left (grey points), shows the sky positions of stars that pass this blue cut in the region around the young cluster.
The larger solid-line circle indicates the region we define as the cluster area, and the two smaller dashed-line circles indicate control fields that are combined and used for modeling the background distribution of proper motions.
The control fields are designed to, together, have the same total area as the cluster field, and were chosen to have similar latitudes as the cluster field in the Magellanic stream coordinate system \citep{Nidever:2008}.
We assume that the background density in the cluster field is equivalent to the average background density of the join control fields.
From visual inspection of the CMDs of the control fields, we do not see any significant clustered over-density and therefore assume that these fields are dominated by the background stellar density.
The right panel of \figurename~\ref{fig:pm-members} shows the difference of the 2D proper motion distributions in the cluster and summed control fields.
The distinct over-density of stars near $(\mu_\alpha, \mu_\delta) \approx (-0.5, 0.5)~\masyr$ is the identified comoving association of blue stars.\footnote{Throughout this article, we use $\mu_\alpha$ to refer to the proper motion value provided by \gaia, which includes the $\cos\delta$ term.}
Note that \figurename~\ref{fig:pm-members} is only meant as an illustration: The cluster proper motion is determined probabilistically by taking into account the full \gaia\ covariance matrices for each source, as described below.

@@

This is also apparent in a position-velocity diagram of the same range of longitudes:
\figurename~\ref{fig:gasspv} shows the velocity of gas in this region (all longitudes, but near $B_{\rm MS} \sim -13^\circ$) as a function of $L_{\rm MS}$.
Here, the longitude of \clustername\ is marked as the vertical (red) line.
The LA II feature is again visible as the $V_{\textrm{LSR}} > 200~\kms$ filament of gas at $L_{\rm MS} \lesssim 60^\circ$ (and a faint extension near $L_{\rm MS} \sim 65^\circ$), but two other prominent filaments with similar morphologies are apparent at lower velocities $V_{\textrm{LSR}} \sim 60~\kms$ and $V_{\textrm{LSR}} \sim 100~\kms$ from $L_{\rm MS} \lesssim 65^\circ$.

@@

\begin{figure*}[ht!]
\centering
\includegraphics[width=0.9\textwidth]{figures/orbits.pdf}
\caption{The past orbit of \clustername in Galactocentric Cartesian coordinates, computed by assuming a line-of-sight velocity equal to each of three different possible gas  filaments in the LA II region of the MS (as indicated in right panel).
In these coordinates, the Sun is at $(x, y, z) = (-8.1, 0, 0)~\kpc$, and the present-day locations of \clustername\ (filled black marker) and the LMC (open black marker) are indicated.
The time at each point in the orbits is indicated by the colorbar, with the darkest regions corresponding to $t = -130~\textrm{Myr}$, the inferred age of \clustername.
The past and future orbit of the LMC are shown as the colored line and dashed line, respectively.
}
\label{fig:orbits}
\end{figure*}

Since the gas distribution in \figurename~\ref{fig:gasspv} shows excess at velocities $V_{\textrm{LSR}} \approx (60, 110, 230)~\km$ near the location of the cluster (ignoring the disk gas between $V_{\textrm{LSR}} \sim -20$--$20~\kms$), it seems that there are three qualitatively different line-of-sight velocities that the gas (and plausibly \clustername) could have.
%: $V_{\textrm{LSR}} \approx (60, 110, 230)~\kms$.
If we adopt these velocity measurements as the line-of-sight (LOS) velocity of \clustername, we have measurements of all 6 phase-space components and can compute orbits for the cluster.
We compute Galactic orbits of the cluster using each of the three values of the line-of-sight velocity from the LA II region filaments.
\figurename~\ref{fig:orbits} shows these three orbits computed by integrating the position of \clustername\ backwards in time:
The two panels of \figurename~\ref{fig:orbits} show the Galactocentric Cartesian trajectories of \clustername\ over the last $260~\textrm{Myr}$ by assuming each of the three plausible line-of-sight velocities discussed above (as labeled in the right panel).
To compute the orbits, we use a three-component Milky Way mass model consisting of a disk \citep{Miyamoto:1975}, a bulge \citep{Hernquist:1990}, and a spherical dark matter halo \citep{Navarro:1996}.
The parameters of this model are set to match the circular velocity profile and disk properties of \citet{Bovy:2015}, and is implemented in the \texttt{gala} package \citep{gala} as the \texttt{MilkyWayPotential}.

@@

We have identified a young, metal-poor stellar association in the Galactic halo --- named \clustername\ --- with an age $\tau \approx \clage$, heliocentric distance $D \approx \cldist$, and metallicity $\feh \approx \clfeh$.
At its present-day sky position, and at large distances, all significant quantities of \hi are associated with the LA and thus it plausibly formed from MS gas.
The age of the cluster is broadly consistent with the time it would have most recently crossed the Galactic midplane, suggesting the possibility that interaction with the Milky Way disk or tidal compression could have triggered this star formation event.
Previous studies have detected young stars in the LA and in the periphery of the LMC \citep{Casetti-Dinescu:2014, Moni-Bidin:2017}, however, this is the first time that an entire young star cluster has been detected so far.% from the Clouds.

The discovery of \clustername\ provides a critical distance constraint to the LA and will aid future Magellanic system and Milky Way modeling efforts.
It also provides an opportunity to study star formation in a unique environment (i.e. low gas density, low metallicity, and large velocity), unlike that of the Milky Way disk or any other local cluster-forming region.
The serendipitous discovery of this cluster is a reminder that the combined value of the \gaia\ data with deep, large-area imaging surveys provides a wealth of information about our Galaxy and stellar halo.

>>>>

We report the discovery of a young ($\tau \sim \clage$), low-mass ($M \sim 1200~\msun$), metal-poor ($\feh \sim \clfeh$) stellar association in the Milky Way halo.
From isochrone fitting to follow-up \decam\ imaging of the cluster, we also measure its distance, $D \approx \cldist$.
At its present Galactocentric position $(R, z) \sim (23, 15)~\textrm{kpc}$, the association is (on the sky) near the leading arm of the gas stream emanating from the Magellanic cloud system, but is located $\approx 60^\circ$ from the Large Magellanic Cloud (LMC) center on the other side of the Milky Way disk.
If we assume that the cluster is co-located with \hi gas in the stream, we directly measure the distance to the leading arm of the Magellanic stream.
At this location relative to the LMC, the measured distance is inconsistent with predictions from models of the LMC/SMC interaction and infall into the Milky Way that do not account for ram pressure and gas interaction with Milky Way disk.
The estimated age of the cluster is consistent with the time of last passage of the leading arm gas through the Galactic midplane, and we therefore speculate that this star-formation event was triggered by the last disk midplane passage at a Galactocentric radius $R \approx 30~\kpc$.
However, most details of this idea remain a mystery: the Magellanic stream has low column density, the Milky Way disk at this large radius has low gas density, and the relative velocity of the leading arm gas and Milky Way gas is large. However it formed, the discovery of a young stellar cluster in the Milky Way halo presents an interesting opportunity for study.
\end{abstract}

@@

We then query all objects from the \gaia\ \DR{2} catalog within a rectangle centered on the nominal position of this group, with a width of $5^\circ$ and a height of $5^\circ$ in the equatorial (ICRS) coordinate system (see Appendix~\ref{sec:queries} for the database query).
\figurename~\ref{fig:cmds} shows the \gaia\ data for this region:
The left panel shows the \gaia\ color-magnitude diagram, extinction corrected following the procedure used in \citet{Danielski:2018}, with the blue over-density highlighted by the polygon (blue) and under-plotted with a $100~\textrm{Myr}$, $\feh = -1.1$ \acronym{MIST} isochrone \citep[red line;][]{Dotter:2016, Choi:2016, Paxton:2011, Paxton:2013, Paxton:2015}.
Surprisingly, this group appears to be a young, distant main sequence, rather than an old population of horizontal branch stars.
The middle and right panels of \figurename~\ref{fig:cmds} show sky positions and proper motions of all stars in this sky region (grey background density), and only stars in the CMD selection polygon (black markers).

The \gaia\ data reveal the presence of a young, distant, spatially-clustered, and co-moving stellar over-density --- named \clustername\ --- but the \gaia\ photometry is too shallow to resolve anything but the brightest main sequence stars.
The spatial morphology of the cluster is large on the sky, and interestingly substructured with at least two subcomponents (labeled a and b in \figurename~\ref{fig:cmds}) that are indistinguishable in terms of their proper motions and uncertainties.
With the \gaia\ data alone, the distance, age, and metallicity of the cluster  cannot be determined, as these quantities are degenerate where the main sequence is nearly vertical.
In the next section, we describe deeper \decam\ imaging obtained over a portion of the cluster.

@@

\figurename~\ref{fig:decam-field} shows the sky positions of the \gaia\ CMD-selected sources (black markers; see same in middle panel, \figurename~\ref{fig:cmds}), along with point sources identified in the $g$-band data obtained with \decam: sources in control fields are shown as red markers, and sources in cluster fields are shown as blue markers.
We note that the smaller over-density located northeast of the \decam\ field, component b (see \figurename~\ref{fig:cmds}), was not followed up in this work.

@@

We measure the mean proper motion of \clustername\ by constructing a probabilistic model of the cluster and background populations using astrometric data from \gaia.
We start by selecting all stars with $(\bprp)_0 < 0.35$ to remove low-mass and old main sequence star contamination in the region.
\figurename~\ref{fig:pm-members}, left (grey points), shows the sky positions of stars that pass this blue cut in the region around the young cluster.
The larger solid-line circle indicates the region we define as the cluster area, and the two smaller dashed-line circles indicate control fields that are combined and used for modeling the background distribution of proper motions.
The control fields are designed to, together, have the same total area as the cluster field, and were chosen to have similar latitudes as the cluster field in the Magellanic stream coordinate system \citep{Nidever:2008}.
We assume that the background density in the cluster field is equivalent to the average background density of the joint control fields.
From visual inspection of the CMDs of the control fields, we do not see any significant clustered over-density and therefore assume that these fields are dominated by the background stellar density.
The right panel of \figurename~\ref{fig:pm-members} shows the difference of the 2D proper motion distributions in the cluster and summed control fields.
The distinct over-density of stars near $(\mu_\alpha, \mu_\delta) \approx (-0.5, 0.5)~\masyr$ is the identified comoving association of blue stars.\footnote{Throughout this article, we use $\mu_\alpha$ to refer to the proper motion value provided by \gaia, which includes the $\cos\delta$ term.}
Note that \figurename~\ref{fig:pm-members} is only meant as an illustration: The cluster proper motion is determined probabilistically by taking into account the full \gaia\ covariance matrices for each source, as described below.

@@

This is also apparent in a position-velocity diagram of the same range of longitudes:
\figurename~\ref{fig:gasspv} shows the velocity of gas in this region (all longitudes, but near $B_{\rm MS} \sim -13^\circ$) as a function of $L_{\rm MS}$.
Here, the longitude of \clustername\ is marked as the dashed vertical (green) line.
The LA II feature is again visible as the $V_{\textrm{LSR}} > 200~\kms$ filament of gas at $L_{\rm MS} \lesssim 60^\circ$ (and a faint extension near $L_{\rm MS} \sim 65^\circ$), but two other prominent filaments with similar morphologies are apparent at lower velocities $V_{\textrm{LSR}} \sim 60~\kms$ and $V_{\textrm{LSR}} \sim 100~\kms$ from $L_{\rm MS} \lesssim 65^\circ$.

\begin{figure*}[t!]
\centering
\includegraphics[width=0.9\textwidth]{figures/orbits.pdf}
\caption{The past orbit of \clustername in Galactocentric Cartesian coordinates, computed by assuming a line-of-sight velocity equal to each of three different possible gas  filaments in the LA II region of the MS (as indicated in right panel).
In these coordinates, the Sun is at $(x, y, z) = (-8.1, 0, 0)~\kpc$, and the present-day locations of \clustername\ (filled black marker) and the LMC (open black marker) are indicated.
The time at each point in the orbits is indicated by the colorbar, with the darkest regions corresponding to $t = -130~\textrm{Myr}$, the inferred age of \clustername.
The past and future orbit of the LMC are shown as the colored line and dashed line, respectively.
}
\label{fig:orbits}
\end{figure*}

@@

Since the gas distribution in \figurename~\ref{fig:gasspv} shows excess at velocities $V_{\textrm{LSR}} \approx (60, 110, 230)~\kms$ near the location of the cluster (ignoring the disk gas between $V_{\textrm{LSR}} \sim -20$--$20~\kms$), it seems that there are three qualitatively different line-of-sight velocities that the gas (and plausibly \clustername) could have.
%: $V_{\textrm{LSR}} \approx (60, 110, 230)~\kms$.
If we adopt these velocity measurements as the line-of-sight (LOS) velocity of \clustername, we have measurements of all 6 phase-space components and can compute orbits for the cluster.
We compute Galactic orbits of the cluster using each of the three values of the line-of-sight velocity from the LA II region filaments.
\figurename~\ref{fig:orbits} shows these three orbits computed by integrating the position of \clustername\ backwards in time:
The two panels of \figurename~\ref{fig:orbits} show the Galactocentric Cartesian trajectories of \clustername\ over the last $260~\textrm{Myr}$ by assuming each of the three plausible line-of-sight velocities discussed above (as labeled in the right panel).
To compute the orbits, we use a three-component Milky Way mass model consisting of a disk \citep{Miyamoto:1975}, a bulge \citep{Hernquist:1990}, and a spherical dark matter halo \citep{Navarro:1996}.
The parameters of this model are set to match the circular velocity profile and disk properties of \citet{Bovy:2015}, and is implemented in the \texttt{gala} package \citep{gala} as the \texttt{MilkyWayPotential}.

@@

We have identified a young, metal-poor stellar association in the Galactic halo --- named \clustername\ --- with an age $\tau \approx \clage$, heliocentric distance $D \approx \cldist$, and metallicity $\feh \approx \clfeh$.
At its present-day sky position, and at large distances, all significant quantities of \hi are associated with the leading arm of the Magellanic stream, and thus it plausibly formed from this gas.
The age of the cluster is broadly consistent with the time it would have most recently crossed the Galactic midplane, suggesting the possibility that interaction with the Milky Way disk or tidal compression could have triggered this star formation event.
Previous studies have detected young stars in the leading arm and in the periphery of the LMC \citep{Casetti-Dinescu:2014, Moni-Bidin:2017}, however, this is the first time that an entire young star cluster has been detected so far from the Clouds.

The discovery of \clustername\ provides a critical distance constraint to the leading arm Magellanic stream and will aid future Magellanic system and Milky Way modeling efforts.
It also provides an opportunity to study star formation in a unique environment (i.e. low gas density, low metallicity, and large velocity), unlike that of the Milky Way disk or any other local cluster-forming region.
The serendipitous discovery of this cluster is a reminder that the combined value of the \gaia\ data with deep, large-area imaging surveys provides a wealth of information about our Galaxy and stellar halo.
